\myTitle

### 2014-08-08 tex/abstract.tex
Incorporate Kristen's comments on manuscript.

      We show that the the algorithm in the program {\sc flock} \citep{Duc&Tur2009} can be
interpreted as an estimation procedure using  
a model that is essentially identical to the {\sc structure} 
\citep{Pritchardetal2000} model with no admixture and non-correlated 
allele frequency priors. The computational routine in {\sc flock} is 
equivalent to searching for the maximum-{\em a-posteriori} (MAP) 
estimate of this restricted {\sc structure} model via a simulated 
annealing algorithm with an extreme cooling 
schedule (namely, the exponent on the objective function~$\rightarrow \infty$).  By analysis of a large, real data set we
confirm that {\sc flock} yields results similar to {\sc structure} with the no-admxixture
model and uncorrelated allele frequency prior, albeit with apparently more variability 
from run to run. Interpreting {\sc flock}'s algorithm as a special case of the model in 
{\sc structure} should aid in  understanding the program's output and behavior.

>>>>

      We show that the the algorithm in the program {\sc flock} \citep{Duc&Tur2009} can be
interpreted as an estimation procedure using  
a model that is essentially identical to the {\sc structure} 
\citep{Pritchardetal2000} model with no admixture and non-correlated 
allele frequency priors. The computational routine in {\sc flock} is 
equivalent to searching for the maximum-{\em a-posteriori} (MAP) 
estimate of this restricted {\sc structure} model via a simulated 
annealing algorithm with an extreme cooling 
schedule (namely, the exponent on the objective function~$\rightarrow \infty$).  By analysis of a large, real data set we
confirm that {\sc flock} yields results similar to {\sc structure} with the no-admxixture
model and uncorrelated allele frequency prior with more variability 
from run to run. Interpreting {\sc flock}'s algorithm as a special case of the model in 
{\sc structure} should aid in  understanding the program's output and behavior.

### 2014-08-08 tex/main-body-text.tex
Incorporate Kristen's comments on manuscript.

Recently, several colleagues suggested we try using the software {\sc flock} \citep{Duc&Tur2009},
for analyzing a number of the data sets from our laboratories.
{\sc flock} is billed as a ``non-Bayesian method [that] 
differs substantially from previous 
clustering algorithms'' \citep[][p.~1333]{Duc&Tur2009}. As described
in a later paper, {\sc flock} ``does not sample the space of partitions through small 
random step walks as in MCMC, and it does not try to optimize some 
target function, such as HWLE\@. Briefly stated, it is not based on a 
probabilistic search algorithm. On the contrary, {\sc flock} is 
entirely deterministic'' \citep[][p.~736]{Duc&Tur2012}.

In trying to understand the behavior of {\sc flock} we identified 
a close relationship between {\sc flock}'s algorithm and the program
{\sc structure}.  In other words, although the algorithm in {\sc flock}
is, as the authors point out, entirely deterministic, it can be interpreted as a 
limiting case of a probabilistic search algorithm applied to one of {\sc structure}'s
models.  More specifically, the {\sc flock} 
algorithm is a special case of the simulated annealing
algorithm for finding the Bayesian maximum-{\em a-posteriori}
estimate from a marginalized form of the {\sc structure} model with no
admixture and non-correlated allele frequencies.

@@

\begin{figure*}
\begin{center}
	 % NOTE. Eric made Flock-Fig1.jpg as a Preview because Flock-Fig1.pdf
	 % ate up too many resources while editing the document.  For final 
	 % production, it should be changed back to Flock-Fig1.pdf
    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1-GreyScale.pdf} % Flock-Fig1.jpg}  
    \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values 
from {\sc flock} made with the program {\sc distruct}. Each horizontal plot represents an individual run of the program {\sc structure}, or the `best run' from a series of runs in {\sc flock}. Within each plot 
$q_i$ values from {\sc structure} and the normalized likelihood values 
from {\sc flock}  are represented by a 
gray-scale vertical bar. The numbers in parentheses on the right indicate how many {\sc structure} runs (out of 6, as indicated) reached the same visually indistinguishable solution, or how many groups of {\sc flock} runs (out of 6) had the same `best run.'}
    \label{fig:qplots}
\end{center}
\end{figure*}

@@

\section*{Conclusions}
In the past decade, the use of genetic markers to identify population structure and to attempt inference of
the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, 
epidemiology, and conservation
genetics. The most commonly used program for doing this is {\sc structure} \citep{Pritchardetal2000,Falushetal2003}, 
a full-featured program with many options that is widely-used, well-tested, and reasonably well-understood. 
We've shown here that {\sc flock}, a program that uses a deterministic algorithm, can be regarded
as providing approximate, maximum-{\em a-posteriori} inference from one of the models implemented in 
{\sc structure}. Given the similarities between the two methods, we expect that {\sc flock}
and {\sc structure} (using the no-admixture model with non-correlated allele frequencies) will deliver
similar results.  We observed such similarity in comparing results from {\sc flock} and {\sc structure} 
on a real data set from \textit{O. mykiss}

Though we observed similar results
between the two programs at low values of $K$, {\sc flock}'s ``best-runs'' solutions became somewhat inconsistent at higher values of $K$, though most solutions were still similar to {\sc structure}'s.
This is likely a consequence of the rapidly increasing size of the search space over possible 
partitions as $K$ and $n$, the number of individuals, increase. 
The likelihood surface defined on partitions of this space will often
be multimodal---possessing numerous local peaks and troughs.
The peak that  {\sc flock}'s deterministic algorithm finds is entirely determined by the initial allocation of 
individuals, as it has no way to visit areas of the surface with smaller likelihood values.
If the number of initial allocations does not increase
as $K$ is increased, the starting allocations become diffusely scattered in the space of all
possible partitions making it less likely that many (if any at all) will converge to the same local mode. This coupled with the fact that there is 
ambiguity in what defines a `best run' when all plateaus are of length zero complicates the interpretation
of results from {\sc flock} if no non-zero plateaus are encountered. Of course, the problem of getting caught in local modes is not unique to {\sc flock}---{\sc structure} itself clearly can wind up sampling predominantly from a fairly isolated part of the space, as evidenced by the different results obtained from different runs of {\sc structure}.

One of the advantages of {\sc flock} over other methods for genetic clustering
is reported to be the short processing time for each run of {\sc flock}.
While \citet{Duc&Tur2012} observed faster processing times (relative to other methods) in their 
comparisons, we did not observe this advantage in our analysis. This appears to result from two issues. First,
our example data set was an order of magnitude larger than the largest data set analyzed in \citet{Duc&Tur2012}.  
Second, in the comparison by \citet{Duc&Tur2012}, 
{\sc flock} was set to perform 50 runs and 20 iterations for each $K$ starting at 2 and ending when 
one of the stopping conditions was met, and computing times for that analysis were compared to ten iterations of {\sc structure} using 
a 50,000 sweep burn-in followed 
by a 200,000 sweep sample from the posterior for values of $K$ from 1 to three more than the true 
value of $K$. 
To us it seemed reasonable to compare, for each $K$, the running times of one iteration of {\sc structure} (50,000 
burn in and 200,000 sample collection sweeps)
to one iteration of {\sc flock} (50 different starting allocations that are reallocated 20 times) 
as both were used to provide one of 6 total estimates (per value of $K$) of the partition of individuals into 
clusters. In short,
it is difficult to reliably assess differences in program run times when there is no clear consensus on what 
constitutes a reasonable comparison with respect to run length and number of runs.

While there has been lengthy 
discourse over the choice of estimators for $K$ in {\sc structure} 
\citep{Pritchardetal2000,Evannoetal2005,Wap&Gag2006,Gaoetal2011}, less has been written on how the number of 
individuals
and the number of starting allocations can affect not only the convergence of different runs of {\sc flock}, but also its 
inference of $K$. \citeauthor{Duc&Tur2012} developed estimation rules for $K$ based on simulated data 
sets with a modest number of individuals and different migration regimes and rates, numbers of loci, and $K$. 
By relying on the number of identical partitions (mean LLOD values) obtained by {\sc flock} over multiple
runs as a means of supporting 
a particular $K$,  similar---but not identical---partitions
will provide no support to a given $K$. If using {\sc flock}, we advise running
multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
{\sc distruct} \citep{Rosenberg2004} to visually inspect the `best run' results, to determine
if there are many similar-but-not-identical partitions found amongst {\sc flock}'s replicate 
solutions.

We should point out that our comparison was between {\sc flock} and the {\em no-admixture} model in 
{\sc structure} which is not the default model used in {\sc structure}. Rather, by default, {\sc structure} uses 
its {\em with-admixture} model.  Such a model explicitly tries to account for the admixed origin
of individuals in the sample and thus may be a more appropriate model than the no-admixture model
(or {\sc flock}) when the sample contains individuals that are admixed between populations.  A further advantage
of the with-admixture model is that it allows inferences for the population of origin of individual alleles at 
different loci within individuals, and not just of the individuals themselves.   We also note that {\sc flock} 
corresponds to {\sc structure}'s model with uncorrelated allele frequency priors. \citet{Falushetal2003} found that
using their ``$F$-model,'' which assumes that allele frequencies between populations are correlated {\em a priori}, 
allowed {\sc structure} to dissect more subtle population structure than was possible using the prior with
uncorrelated allele frequencies.  Modifying {\sc flock} so as to reap similar benefits from using a prior
with correlated allele frequencies seems like it would be an interesting challenge, though difficult
because {\sc flock}'s formulation effectively integrates the allele frequencies out of the model.

>>>>

The recently introduced software {\sc flock} \citep{Duc&Tur2009} has
been described
as a ``non-Bayesian method [that] 
differs substantially from previous 
clustering algorithms'' \citep[][p.~1333]{Duc&Tur2009}. As reported
in a later paper, ``{\sc flock} is very different from  other clustering 
programs. It does not sample the space of partitions through small 
random step walks as in MCMC, and it does not try to optimize some 
target function, such as HWLE\@. Briefly stated, it is not based on a 
probabilistic search algorithm. On the contrary, {\sc flock} is 
entirely deterministic'' \citep[][p.~736]{Duc&Tur2012}.

However, here we identify 
the close relationship between {\sc flock}'s algorithm and the program
{\sc structure}.  Although the algorithm in {\sc flock}
is, as the authors point out, entirely deterministic, it can be interpreted as a 
limiting case of a probabilistic search algorithm applied to one of {\sc structure}'s
models.  More specifically, the {\sc flock} 
algorithm is a special case of the simulated annealing
algorithm for finding the Bayesian maximum-{\em a-posteriori}
estimate from a marginalized form of the {\sc structure} model with no
admixture and non-correlated allele frequencies.

@@

\begin{figure*}
\begin{center}
    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1.pdf}   % Use this if we want it grayscale -> Flock-Fig1-GreyScale.pdf  
    \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values 
from {\sc flock} made with the program {\sc distruct}. Each horizontal plot represents an individual run of the program {\sc structure}, or the `best run' from a series of runs in {\sc flock}. Within each plot 
$q_i$ values from {\sc structure} and the normalized likelihood values 
from {\sc flock}  are represented by a 
gray-scale vertical bar. The numbers in parentheses on the right indicate how many {\sc structure} runs (out of 6, as indicated) reached the same visually indistinguishable solution, or how many groups of {\sc flock} runs (out of 6) had the same `best run.'}
    \label{fig:qplots}
\end{center}
\end{figure*}

@@

\section*{Conclusions}
In the past decade, the use of genetic markers to identify population structure and to attempt inference of
the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, 
epidemiology, and conservation
genetics. The most commonly used program for doing this is {\sc structure} \citep{Pritchardetal2000,Falushetal2003}, 
a full-featured program with many options that is widely-used, well-tested, and universally regarded as a standard tool amongst molecular ecologists. 
We've shown here that {\sc flock} is essentially equivalent to the program {\sc structure} using the no-admixture model with non-correlated allele frequencies. In comparing results from {\sc flock} and {\sc structure} 
on a real data set from {\em O. mykiss} we find that the two programs give similar results, only {\sc flock}'s results at higher values of $K$ appear to be more variable.

Though we observed similar results
between the two programs at low values of $K$, {\sc flock}'s `best-run' solutions became somewhat inconsistent at higher values of $K$.
This is likely a consequence of the rapidly increasing size of the search space over possible 
partitions as $K$ and $n$, the number of individuals, increase. 
The likelihood surface defined on partitions of this space will often
be multimodal---possessing numerous local peaks and troughs.
The peak that  {\sc flock}'s deterministic algorithm finds is entirely determined by the initial allocation of 
individuals, as it is a deterministic algorithm.
If the number of initial allocations does not increase
as $K$ is increased, the starting allocations become diffusely scattered in the space of all
possible partitions making it less likely that many (if any at all) will converge to the same local mode. This coupled with the fact that there is 
ambiguity in what defines a `best run' when all plateaus are of length zero complicates the interpretation
of results from {\sc flock} if no non-zero plateaus are encountered. The problem of getting 
caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flock}.

One of the advantages of {\sc flock} over other methods for genetic clustering
is reported to be the short processing time for each run of {\sc flock}.
While \citet{Duc&Tur2012} observed faster processing times (relative to other methods) in their 
comparisons, we did not observe this advantage in our analysis. This appears to result from two issues. First,
our example data set was an order of magnitude larger than the largest data set analyzed in \citet{Duc&Tur2012}.  
Second, in the comparison by \citet{Duc&Tur2012}, 
{\sc flock} was set to perform 50 runs and 20 iterations for each $K$ starting at 2 and ending when 
one of the stopping conditions was met, and computing times for that analysis were compared to ten iterations of {\sc structure} using 
a 50,000 sweep burn-in followed 
by a 200,000 sweep sample from the posterior for values of $K$ from 1 to three more than the true 
value of $K$. 
To us it seemed reasonable to compare, for each $K$, the running times of one iteration of {\sc structure} (50,000 
burn in and 200,000 sample collection sweeps)
to one iteration of {\sc flock} (50 different starting allocations that are reallocated 20 times) 
as both were used to provide one of 6 total estimates (per value of $K$) of the partition of individuals into 
clusters. Though
it is difficult to reliably assess differences in program run times when there is no clear consensus on what 
constitutes a reasonable comparison with respect to run length and number of runs, {\sc flock} did not carry a distinct run-time advantage over {\sc structure} in our analyses.

While there has been lengthy 
discourse over the choice of estimators for $K$ in {\sc structure} 
\citep{Pritchardetal2000,Evannoetal2005,Wap&Gag2006,Gaoetal2011}, less has been written on how the number of 
individuals
and the number of starting allocations can affect not only the convergence of different runs of {\sc flock}, but also its 
inference of $K$. \citeauthor{Duc&Tur2012} developed estimation rules for $K$ based on simulated data 
sets with a modest number of individuals and different migration regimes and rates, numbers of loci, and $K$. 
By relying on the number of identical partitions (mean LLOD values) obtained by {\sc flock} over multiple
runs as a means of supporting 
a particular $K$,  similar---but not identical---partitions
will provide no support to a given $K$. Given {\sc flock}'s tendency to converge to 
different solutions with large $K$ and $n$, it seems that 
this method for estimating $K$ is unlikely to be reliable or workable in large, complex
problems.  One possible patch for this problem would be running
multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
{\sc distruct} \citep{Rosenberg2004} to visually inspect the `best run' results, and then 
determine plateaus amongst {\sc flock}'s replicate solutions not by identical LLOD, but by visual inspection of 
partitions that
are very similar but are perhaps not identical; however such an approach would be labor-intensive
and would rely on arbitrary, difficult-to-reproduce decisions.  It might be better to use {\sc structure} in such cases.
Estimating $K$ is a hard problem---especially in real populations that seldom conform to the assumptions of genetic clustering models---and should be undertaken with caution, regardless of the method 
used.

We should point out that our comparison was between {\sc flock} and the {\em no-admixture} model in 
{\sc structure} which is not the default model used in {\sc structure}. Rather, by default, {\sc structure} uses 
its {\em with-admixture} model.  Such a model explicitly tries to account for the admixed origin
of individuals in the sample and thus may be a more appropriate model than the no-admixture model,
or {\sc flock}, when the sample contains individuals that are admixed between populations.  A further advantage
of the with-admixture model is that it allows inferences for the population of origin of individual alleles at 
different loci within individuals, and not just of the individuals themselves.   We also note that {\sc flock} 
corresponds to {\sc structure}'s model with uncorrelated allele frequency priors. \citet{Falushetal2003} found that
using their ``$F$-model,'' which assumes that allele frequencies between populations are correlated {\em a priori}, 
allowed {\sc structure} to dissect more subtle population structure than was possible using the prior with
uncorrelated allele frequencies.  Modifying {\sc flock} so as to reap similar benefits from using a prior
with correlated allele frequencies seems like it would be an interesting challenge, though difficult
because {\sc flock}'s formulation effectively integrates the allele frequencies out of the model.
